{\bf {\Large Title....}}

### 2016-07-01 ident_manu.Rnw
identifiability by category plot

%knitr options
<<setup,include=F,cache=F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, width = 90, digits = 1)

source('R/funcs.r')
library(WRTDStidal)
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ncdf4)
@

@@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton (up to six groups) and zooplankton (two groups) as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, a set of parameter values for each of the equations is also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert-based knowledge in absence of the former.  The sensitivity of \ac{do} was estimated at each timestep from FishTank in relation to 249 parameters that are included in the 108 equations.  For simplicity in the discussion below, the parameters were grouped into one of six categories in relation to applicable equations: optics, phytoplankton, zooplankton, temperature, and organic matter.  A full description of the equations and parameters is available as an appendix in \citepalias{LehrterIR}.  
\subsection{Local sensitivity analysis}

@@

In general, positive sensitivity estimates suggested a parameter had a positive effect on \ac{do} for a given increase in the parameter, whereas the converse was true for negative sensitivity estimates.  However, the effect of a parameter change may not be uniform over time such that $S_{ij}$ can change in magnitude and sign depending on the location.  Time series of \ac{do} estimates before and after perturbation were also evaluated to identify patterns not captured by the summary statistics. All parameters for each of the six equation categories (optics, phytoplankton, zooplankton, temperature, and organic matter) that had non-zero $L1$ or $L2$ were retained for identifiability analysis.

@@

\begin{equation}
\gamma = \frac{1} {\sqrt{ \min \left(\rm{EV}[\mathit{\hat{S}}^\top \mathit{\hat{S}}]\right)}}
\end{equation}

@@

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above.  These rulesets were developed with the assumption that parameters will be selected given preference for those with high sensitivity and identifability based on $\gamma < 15$ was an acceptable threshold for subsets (e.g., 93\% accountability between parameters).  Selection heurestics also recognized that parameter categories (i.e., optics, phytoplankton, zooplankton, temperature, organic matter) may have unequal preferences given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item independent of parameter category,
\item blocked within parameter category,
\item or considering all categories equally
\end{inparaenum}.  The selection rules produced seven subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

\cref{tab:optsens,tab:tempsens,tab:phytosens,tab:zoopsens,tab:omsens}

@@

The parameters that have the greatest effect on the model by category (optics, organics, phytoplankton, zooplankton) are as follows.

Plotting the raw values from the sensitivity analysis provides a visual assessment of changes.

@@

% combination example
<<combnex, echo = F, fig.cap = 'Number of unique combinations that can occur by selecting parameters for optimization from different parameter sets.  The number of combinations are shown for increasing numbers of selected parameters from the total in the set, where 100 parameter sets are shown each with one through 100 total parameters.  For example, there are six unique parameter combinations that result from choosing two parameters from a set.  Note that the number of unique combinations is shown as the natural-log.  The same information is viewed differently in each plot.', fig.height = 5, fig.width = 5, out.width = '0.6\\textwidth'>>=

@@

% sensitivity example of top
<<sensplo, fig.cap = 'Sensitivity of \\ac{do} to parameter changes.  Each solid line shows the parameter within each category that had the largest effect on \\ac{do} with a 50\\% change from the default value.  The dashed line shows \\ac{do} holding all parameters at their default values.  The bottom plot shows the estimated \\ac{do} as a difference from the default.', fig.height = 5.5, fig.width = 8, echo = F, cache = TRUE>>=
load(file = 'data/sens_ests.RData')
load(file = 'data/sens_ests_cat.RData')

@@

toplo <- filter(sens_ests$raw, Parameter %in% toeval) %>% 
  mutate(
    diffest = Estimate - default,
    time = as.POSIXct(time, origin = as.POSIXct('2002-01-01 0:0', tz = "GMT"), tz = 'GMT'),
    time = as.Date(time)
  )

@@

# align widths of plots in first column, first two
pA <- ggplot_gtable(ggplot_build(p1))
pB <- ggplot_gtable(ggplot_build(p2))
maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3])

# Set the widths
pA$widths[2:3] <- maxWidth
pB$widths[2:3] <- maxWidth

grid.arrange(pA, pB, ncol = 1, heights = c(1, 0.85))
@

>>>>

%knitr options
<<setup,include=F,cache=F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.path = 'figs/', fig.align = 'center', fig.show = 'hold', message = F, echo = F, results = 'asis', dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign = TRUE, width = 90)

source('R/funcs.r')
library(WRTDStidal)
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ggrepel)
@

@@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton (up to six groups) and zooplankton (two groups) as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, a set of parameter values for each of the equations is also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert-based knowledge in absence of the former.  The sensitivity of \ac{do} was estimated at each timestep from FishTank in relation to 249 parameters that are included in the 108 equations.  For simplicity in the discussion below, the parameters were grouped into one of six categories based on respective equations: optics, organic matter, phytoplankton, temperature, and zooplankton.  A full description of the equations and parameters is available as an appendix in \citepalias{LehrterIR}.  
\subsection{Local sensitivity analysis}

@@

In general, positive sensitivity estimates suggested a parameter had a positive effect on \ac{do} for a given increase in the parameter, whereas the converse was true for negative sensitivity estimates.  However, the effect of a parameter change may not be uniform over time such that $S_{ij}$ can change in magnitude and sign depending on the location.  Time series of \ac{do} estimates before and after perturbation were also evaluated to identify patterns not captured by the summary statistics. All parameters for each of the six equation categories (optics, organic matter, phytoplankton, temperature, and zooplankton) that had non-zero $L1$ or $L2$ were retained for identifiability analysis.

@@

\begin{equation} \label{gameq}
\gamma = \frac{1} {\sqrt{ \min \left(\rm{EV}[\mathit{\hat{S}}^\top \mathit{\hat{S}}]\right)}}
\end{equation}

@@

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above.  These rulesets were developed with the assumption that parameters will be selected given preference for those with high sensitivity and identifability based on $\gamma < 15$ was an acceptable threshold for subsets (e.g., 93\% accountability between parameters).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item independent of parameter category,
\item blocked within parameter category,
\item or considering all categories equally
\end{inparaenum}.  The selection rules produced seven subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

@@

\cref{tab:optsens,tab:tempsens,tab:phytosens,tab:zoopsens,tab:omsens}, how many parameters in each category, how many induced a response in \ac{do}, what parameters in each category had the greatest sensitivity (\cref{fig:sensplo}), what were sensitivity magnitudes between categories, did these vary by phyto or zoop groups?

Plotting the raw values from the sensitivity analysis provides a visual assessment of changes (\cref{fig:sensplot}).

@@

% combination example
<<combnex, echo = F, fig.cap = 'Examples of unique parameter combinations from different parameter sets and number of selected parameters.  The number of combinations are shown for increasing numbers of selected parameters from the total in the set, where 50 parameter sets are shown each with one through 50 total parameters. Note that the number of unique combinations is shown as the natural-log.', fig.height = 5, fig.width = 5, out.width = '0.6\\textwidth'>>=

@@

% sensitivity example of top
<<sensplo, fig.cap = 'Sensitivity of \\ac{do} to parameter changes. The solid lines show the change in \\ac{do} based on a 50\\% change from the default parameter values (dashed line) for each parameter.  Individual parameters with the largest effect are shown for each category.  The top plot shows the model output and the middle plot shows the estimated \\ac{do} as a difference from the default.  The bottom plot shows the distribution of error values (as $log\\left(L1\\right)$) for all parmaters in each category.', fig.height = 8, fig.width = 8, echo = F, cache = TRUE>>=
load(file = 'data/sens_ests.RData')
load(file = 'data/sens_ests_cat.RData')

@@

# for top sens plots
toplo <- filter(sens_ests$raw, Parameter %in% toeval) %>% 
  mutate(
    diffest = Estimate - default,
    time = as.POSIXct(time, origin = as.POSIXct('2002-01-01 0:0', tz = "GMT"), tz = 'GMT'),
    time = as.Date(time), 
    Parameter = factor(Parameter, levels = c('astar490_1', 'k11_1', 'Qc_1', 'Tref(nospA+nospZ)_4', 'ZKa_1'))
  )

# for all L1 boxplots
toplo2 <- mutate(sens_ests_cat,
  Category = factor(Category, 
    levels = c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton')
    ), 
  Parameter = as.character(Parameter)
  ) %>% 
  group_by(Category) %>% 
  mutate(
    tops = ifelse(error == max(error), Parameter, NA)
  )

@@

p3 <- ggplot(toplo2, aes(x = Category, y = log(error), fill = Category)) + 
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(palette = colset) + 
  scale_y_continuous(expression(paste('log(', italic('L1'), ')'))) + 
  theme(
    legend.position = 'none',
    axis.title.x = element_blank(),
    axis.line.x = element_line(), 
    axis.line.y = element_line()
    )

# align widths of plots in first column, first two
pA <- ggplot_gtable(ggplot_build(p1))
pB <- ggplot_gtable(ggplot_build(p2))
pC <- ggplot_gtable(ggplot_build(p3))
maxWidth = grid::unit.pmax(pA$widths[2:3], pB$widths[2:3], pC$widths[2:3])

# Set the widths
pA$widths[2:3] <- maxWidth
pB$widths[2:3] <- maxWidth
pC$widths[2:3] <- maxWidth

grid.arrange(pA, pB, pC, ncol = 1, heights = c(1, 0.85, 0.85))
@

% identifiability boxplots
<<identbox, fig.cap = 'Identifiability (as $\\gamma$, \\cref{gameq}) of parameter subsets for \\ac{do}.  Plots in (a) show identifiability by parameter caterogires (b) shows identifiability by seleting the top 1 through 5 parameters regardless of category.  Lines represent identifiability ranges for each the number of parameters in the set.  The phytoplankton category is limited to 25 total parameters.', fig.height = 8, fig.width = 8, echo = F, cache = TRUE>>=
# identifiability by category (all parameters with >0 L1) 
data(cat_ident)
data(top_ident)

colset <- 'Set2'

toplo <- cat_ident %>% 
  lapply(., function(x) dplyr::select(x, N, coll)) %>% 
  reshape2::melt(., id.var = c('N', 'coll')) %>% 
  rename(Category = L1) %>% 
  mutate(Category = factor(Category, 
    levels = c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton'),
    labels = c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton')
    )) %>% 
  filter(N <= 25) %>% # reduced for plot space
  group_by(N, Category) %>% 
  summarise(
    ave = mean(coll), 
    med = median(coll),
    min = min(coll), 
    max = max(coll)
  )

p1 <- ggplot(toplo, aes(x = N, colour = Category, group = N)) + 
  geom_point(aes(y = ave)) + 
  geom_point(aes(y = med)) + 
  geom_errorbar(aes(ymin = min, ymax = max), width = 0) + 
  facet_wrap(~ Category, ncol = 1, scales = 'free_y') +
  theme_minimal() + 
  theme(
    axis.line.x = element_line(), 
    axis.line.y = element_line(), 
    axis.title.x = element_blank(), 
    axis.title.y = element_blank()
  ) + 
  scale_colour_brewer(palette = colset) + 
  theme(legend.position = 'none') + 
  ggtitle('(a) By parameter category')

# identifiability top 1, 2, 3, 4, 5 all categories combined
toplo2 <-  reshape2::melt(top_ident, id.var = c('N', 'coll')) %>% 
  select(N, coll, L1) %>% 
  mutate(L1 = factor(L1, levels = paste0('top', 1:5), labels = paste0('Top ', 1:5))) %>% 
  rename(npercat = L1) %>% 
  group_by(N, npercat) %>% 
  summarise(
    ave = mean(coll), 
    med = median(coll),
    min = min(coll), 
    max = max(coll)
  )

p2 <- ggplot(toplo2, aes(x = N, colour = npercat, group = N)) + 
  geom_point(aes(y = ave, shape = 'Ave')) + 
  geom_point(aes(y = med, shape = 'Med')) + 
  geom_errorbar(aes(ymin = min, ymax = max), width = 0) + 
  facet_wrap(~ npercat, ncol = 1, scales = 'free_y') +
  theme_minimal() + 
  scale_colour_brewer(palette = colset, guide = F) + 
  theme(
    axis.line.x = element_line(), 
    axis.line.y = element_line(), 
    axis.title.x = element_blank(), 
    axis.title.y = element_blank(), 
    legend.title = element_blank()
  ) + 
  ggtitle('(b) By top parameters, all categories')

pleg <- g_legend(p2)
p2 <- p2 + theme(legend.position = 'none')

grid.arrange(
  arrangeGrob(p1, p2, ncol = 2),
  left = 'Identifiability', 
  bottom = 'Parameters in set',
  pleg, ncol = 2, widths = c(1, 0.1)
)
@

### 2016-07-01 ident_manu.tex
identifiability by category plot

\begin{singlespace}
\title{{\bf {\Large Title....}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Wed Jun 29 14:50:46 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

\caption[Number of unique combinations that can occur by selecting parameters for optimization from different parameter sets]{Number of unique combinations that can occur by selecting parameters for optimization from different parameter sets.  The number of combinations are shown for increasing numbers of selected parameters from the total in the set, where 100 parameter sets are shown each with one through 100 total parameters.  For example, there are six unique parameter combinations that result from choosing two parameters from a set.  Note that the number of unique combinations is shown as the natural-log.  The same information is viewed differently in each plot.}\label{fig:combnex}
\end{figure}

@@

\caption{Sensitivity of \ac{do} to parameter changes.  Each solid line shows the parameter within each category that had the largest effect on \ac{do} with a 50\% change from the default value.  The dashed line shows \ac{do} holding all parameters at their default values.  The bottom plot shows the estimated \ac{do} as a difference from the default.}\label{fig:sensplo}
\end{figure}

>>>>

\begin{singlespace}
\title{{\bf {\Large Title....}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Wed Jun 29 16:36:46 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

\caption[Examples of unique parameter combinations from different parameter sets and number of selected parameters]{Examples of unique parameter combinations from different parameter sets and number of selected parameters.  The number of combinations are shown for increasing numbers of selected parameters from the total in the set, where 50 parameter sets are shown each with one through 50 total parameters. Note that the number of unique combinations is shown as the natural-log.}\label{fig:combnex}
\end{figure}

@@

\caption{Sensitivity of \ac{do} to parameter changes. The solid lines show the change in \ac{do} based on a 50\% change from the default parameter values (dashed line) for each parameter.  Individual parameters with the largest effect are shown for each category.  The top plot shows the model output and the middle plot shows the estimated \ac{do} as a difference from the default.  The bottom plot shows the distribution of error values (as $log\left(L1\right)$) for all parmaters in each category.}\label{fig:sensplo}
\end{figure}

% identifiability boxplots
\begin{figure}[!ht]

{\centering \includegraphics[width=\maxwidth]{figs/identbox-1}

\caption[Identifiability (as ]{Identifiability (as $\gamma$, \cref{gameq}) of parameter subsets for \ac{do}.  Plots in (a) show identifiability by parameter caterogires (b) shows identifiability by seleting the top 1 through 5 parameters regardless of category.  Lines represent identifiability ranges for each the number of parameters in the set.  The phytoplankton category is limited to 25 total parameters.}\label{fig:identbox}
\end{figure}
